The Cost of Undisturbed Landscapes

### 2020-05-05 projects/asparagus/doc/paper_clean.tex
updated asparagus paper_clean.tex

% SI units
\usepackage{siunitx}
\DeclareSIUnit{\thermal}{th}
\DeclareSIUnit{\electric}{el}
\DeclareSIUnit{\year}{a}

@@

    \section{Results and Discussion} \label{sec:results-discussion}
    To derive the social opportunity cost of not using wind power (but solar photovoltaics instead), we start by
    determining the long-run power market equilibrium without constraints on renewable capacity addition.
    This gives us our baseline scenario with minimal system costs.
    Starting from this baseline scenario, we evaluate further scenarios in which we gradually tighten the upper limit
    on wind power expansion, for example due to social conflict around wind turbine impacts on landscapes.
    We continue to tighten upper limits on wind power expansion till no wind power is allowed anymore.
    For each step, we observe system cost, total \COO emissions and other variables of interest.
    By relating the change in generated wind energy to the change in system cost (or other variables of interest), we
    derive a measure of the (approximate) marginal impact of wind power expansion (or contraction).

    In the baseline scenario, we assume capacity-weighted average capital cost of solar PV of $630$ \EUR per kW
    installed.
    This corresponds to about 56\% of PV installations being mounted on roof tops, while the remainder (44\%) is
    realized as open-space solar PV at utility scale.
    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to capital cost of solar
    PV\@.
    Moreover, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
    internalized through \COO pricing.
    In the baseline scenario, we presume \COO prices as low as $30$ \EUR per tonne, as there is no incentive to add
    carbon-free generation capacities above the policy target.
    As future \COO prices are uncertain but do impact renewables deployment ~\citep{Brown2020, Kirchner2019}, we
    extend the analysis to \COO prices of $60$, $90$, and $120$ \EUR per tonne.

    \todo[inline]{describe changes in system configuration}

    \todo[inline]{describe marginal change in system cost due to less wind deployed}

    \todo[inline]{insert table with key scenario results}
    \begin{table}
        \caption{Key scenario results}
        \begin{tabulary}{\textwidth}{LLC}
            \toprule
            column & column & column \\
\midrule
AAA & BBB & CCC \\
\bottomrule
\end{tabulary}
\end{table}

\begin{figure*}
\centering
\begin{subfigure}[b]{0.475\textwidth}
\centering
\includegraphics[width=\textwidth]{./figures/undisturbed_base_share.pdf}
\caption{Relative substitution}
\label{fig:rel_substitution}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.475\textwidth}
\centering
\includegraphics[width=\textwidth]{./figures/undisturbed_base.pdf}
\caption{Absolute substitution}
\label{fig::abs_substitution}
\end{subfigure}
\caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
\label{figure:opportunity-cost-base}
\end{figure*}

\subsection{Sensitivity Analysis} \label{subsec:sensitivity}
A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV.
Thus, we complement our baseline scenario with a sensitivity analysis on investment cost.
For this purpose, we vary the investment cost of solar PV between our baseline assumption of 630 EUR/kW and 275
EUR/kW that \cite{Vartiainen2019} project for utility-scale solar PV in the year 2030.

\begin{figure*}
\centering
\includegraphics[width=0.8 \textwidth]{./figures/undisturbed_low.pdf}
\caption{Opportunity cost of wind power}
\end{figure*}

%    Further uncertainty surrounds
%    •	Storage prices
%    •	Weather / full-load hours
%    •	LOOK AT WHAT HAPPENS WHEN GERMANY BUILDS MORE PV THAN ANTICIPATED (particularly in the low pv-cost scenario)

\subsection{Limitations of the Analysis}
To little surprise, the optimal deployment of renewable energy technologies heavily depends on underlying capital
cost.
Alas, uncertainty regarding future capital cost is high.
\todo[inline]{WRITE SOMETHING ON THE IMPACT OF DIFFERENT COST ASSUMPTIONS ON TOTAL SYSTEM COST.}

Nevertheless, several findings are remarkably robust to changing capital cost.
A higher penetration level of PV leads to increasing emissions of CO2 in the least-cost systems we investigated.
Due to the strong seasonal differences in electricity generation from solar PV, systems relying heavily on this technology either need to deploy excess PV capacity to boost generation in winter (resulting in heavy curtailment in summer) or, alternatively, add additional capacity to make up for the winter shortfall in electricity generation from solar PV.
This additional capacity could be storage capacity to transfer summer excesses to winter.
However, storages would have to balance seasonal differences in this case, implying the need for very high storage capacities, which are costly to realize.
Typically, it will be cheapest to add (at least some) fossil-fueled backup plants, even at \COO prices of EUR 70 per ton.
\todo[inline]{HERE, WE COULD INVESTIGATE AT WHICH CO2-PRICE FOSSIL BACKUP VANISHES.}

Moreover, the observed increase in producer surplus in line with higher CO2 penetration is robust to capital cost uncertainty.
Higher levels of PV penetration require more efforts for system integration.
\todo[inline]{THIS CAN'T BE THE FULL EXPLANATION AS PROD SURPLUS DECREASES AT VERY HIGH LEVELS OF PV}

@@

>>>>

% SI units
\usepackage{siunitx}
\DeclareSIUnit{\thermal}{th}
\DeclareSIUnit{\electric}{el}
\DeclareSIUnit{\year}{a}
\DeclareSIUnit{\wattpeak}{Wp}

@@

    \section{Results and Discussion} \label{sec:results-discussion}
    To derive the social opportunity cost of not using wind power (but solar photovoltaics instead), we start by
    determining the long-run power market equilibrium without constraints on renewable capacity addition.
    This gives us our baseline scenario with minimal system costs.
    Starting from this baseline scenario, we evaluate further scenarios in which we gradually tighten the upper limit
    on wind power expansion, for example due to social conflict around wind turbine impacts on landscapes.
    We continue to tighten upper limits on wind power expansion till no wind power is allowed anymore.
    For each step, we observe system cost, total \COO emissions and other variables of interest.
    By relating the change in generated wind energy to the change in system cost (or other variables of interest), we
    derive a measure of the (approximate) marginal impact of wind power expansion (or contraction).
    A detailed discussion of model results is provided in \ref{sec:detailedresults}.

    \begin{figure*}[h!t]
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{./figures/undisturbed_base_share.pdf}
            \caption{Relative substitution}
            \label{fig:rel_substitution}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{./figures/undisturbed_base.pdf}
            \caption{Absolute substitution}
            \label{fig::abs_substitution}
        \end{subfigure}
        \caption{Opportunity cost of wind power assuming PV overnight cost of 630 EUR/kWp}
        \label{figure:opportunity-cost-base}
    \end{figure*}

    With a gradual restriction of deployable wind power capacities, we find an opportunity costs of wind power
    to increase by about $3\,900$ \EUR for each \SI{}{\mega\watt} wind power substituted by solar PV across all scenarios
    that we studied.
    At PV overnight cost of $630$ \EUR per MWp (corresponding to a mix of 55\% rooftop and 45\% open space
    installations), the opportunity cost of the last \SI{}{\mega\watt} wind power replaced reaches close to $62\,500$ \EUR.
    At a wind power capacity consistent with the policy goal of generating \SI{21}{\tera\watt\hour} of electricity
    from renewable sources, opportunity cost for wind power are in close to $20\,000$ \EUR per MW, with a notable
    exception for \COO prices of $60$ \EUR per tonne.
    In this case, opportunity cost of wind power are estimated at close to $10\,000$ \EUR per MW, about half the size of
    opportunity cost estimated for all other \COO price scenarios.
    \todo[inline]{investigate why this happens}
    \todo[inline]{add brief explanation for why opportunity cost increase -- more thermal dispatch \& capacity etc.}

    \subsection{Sensitivity Analysis} \label{subsec:sensitivity}
    A major factor of influence for our results are the assumed capital cost of wind turbines and solar PV\@.
    Therefore, we complement our baseline scenario with a sensitivity analysis on overnight investment cost of solar
    photovoltaics.
    We gradually lower the capital cost of solar PV from our baseline assumption of \SI{630}[\$]{\per\kilo\wattpeak}
    (corresponding to \SI{55}{\percent} of rooftop and \SI{45}{\percent} open space capacity) to
    \EUR~\SI{275}{\per\kilo\wattpeak}, which Vartiainen et al.~\cite{Vartiainen2019} project for open-space,
    utility-scale solar PV in the year 2030.

    \begin{figure*}[h!t]
        \centering
        \includegraphics[width=0.8 \textwidth]{./figures/undisturbed_low.pdf}
        \caption{Opportunity cost of wind power}
    \end{figure*}

    Under these assumptions, the impact of \COO prices on the opportunity cost of wind power increases.
    If all additional PV is realized as open-space, utility-scale installation (i.e.\ PV capital cost of
    \EUR~\SI{275}{\per\kilo\wattpeak}), the optimal deployment of wind power declines and approaches zero at \COO prices
    of \EUR~\SI{30}{\per\tonne} or below.
    In consequence, we can not determine opportunity cost of wind power given that both \COO prices and capital cost of
    solar PV are very low.
    For higher \COO prices, the optimal deployment of wind power is strictly positive, even at low capital cost of solar
    PV and we find opportunity cost to increase by about $2\,550$ \EUR for each MW wind power foregone.
    The last MW of wind power substituted with solar PV comes at opportunity cost between $20\,000$ and $56\,000$ \EUR,
    depending on the \COO price prevailing.

    \subsection{Limitations of the Analysis}
    To little surprise, the optimal deployment of renewable energy technologies heavily depends on underlying capital
    cost.
    Alas, uncertainty regarding future capital cost is high.
    \todo[inline]{write about impact of cost assumptions on total system cost}

    Nevertheless, several findings are remarkably robust to changing capital cost.
    A higher penetration level of PV leads to increasing emissions of CO2 in the least-cost systems we investigated.
    Due to the strong seasonal differences in electricity generation from solar PV, systems relying heavily on this technology either need to deploy excess PV capacity to boost generation in winter (resulting in heavy curtailment in summer) or, alternatively, add additional capacity to make up for the winter shortfall in electricity generation from solar PV.
    This additional capacity could be storage capacity to transfer summer excesses to winter.
    However, storages would have to balance seasonal differences in this case, implying the need for very high storage capacities, which are costly to realize.
    Typically, it will be cheapest to add (at least some) fossil-fueled backup plants, even at \COO prices of EUR 70 per ton.
    \todo[inline]{do we know if there is a co2-price at which fossil backup vanishes? -> at 120 eur / t fossil backup is about halved.}

@@

    \section{Detailed Description of scenario results} \label{sec:detailedresults}

    In the baseline scenario, we assume capacity-weighted average capital cost of solar PV of $630$ \EUR per kW
    installed.
    This corresponds to about 56\% of PV installations being mounted on roof tops, while the remainder (44\%) is
    realized as open-space solar PV at utility scale.
    In section~\ref{subsec:sensitivity} we analyze the sensitivity of our results with respect to capital cost of solar
    PV\@.
    Moreover, we assume efficient pricing of \COO emissions, i.e.\ all otherwise external cost of \COO emissions are
    internalized through \COO pricing.

    Future \COO prices impact renewables deployment ~\citep{Brown2020, Kirchner2019} and affect the results of our
    analysis.
    As we can not foresee \COO prices in the year 2030, we have conducted our analysis for \COO prices in the range of
    $0$ and $120$ \EUR per tonne.
    In this context, it is worth noting that within the framework of our analysis a \COO price of $30$ \EUR per tonne
    would be consistent with RET expansion goals stated by the Austrian government.
    At \COO price above $30$ \EUR per tonne, there is incentive to invest in \COO emission-free generation beyond
    government targets.

    \subsubsection{Changes in system configuration}
    \todo[inline]{describe changes in system configuration}
    Consistent across all \COO price scenarios, the few existing, small oil-fired units in Austria are shut down.
    Conversely, natural gas-fired co-generation capacities are expanded.
    Additions of natural-gas fired capacities are lower at high \COO prices and at high wind power installations.
    However, there are huge differences in fossil capacity additions between \COO prices of $60$ \EUR per ton and $90$
    \EUR per ton.
    In the former case, there is very little difference to capacity addition at zero \COO price, while in the latter case
    fossil capacity additions are more than halved at high wind shares.
    Further heat generation is taking place in newly added compression heat pumps, of which between \SI{1.2}{\giga\watt}
    and \SI{2.9}{\giga\watt} are active in Austria.
    Generally, higher \COO prices are leading to higher installed capacities of heat pumps.
    Similarly, optimally installed heatpump capacities tends to fall as wind power is substituted with solar PV.
    This effect is more pronounced at high \COO prices and lower in magnitude than the level effect of \COO prices.

    \subsubsection{Changes in system operation}
    Curtailment of renewable electricity generation in Austria is the highest when \COO prices are high and the system
    relies on large shares of wind power.
    Curtailment reaches its minimum at intermediate \COO prices of $60$ \EUR per tonne and wind power installations of
    \SI{8}{\giga\watt}.
    At \COO prices of $30$ \EUR per tonne or lower curtailment tends to increase as wind power is substituted by
    photovoltaics.

    \todo[inline]{describe marginal change in system cost due to less wind deployed}
    \todo[inline]{insert table with key scenario results}

    \begin{table}
        \centering
        \caption{Key scenario results}
        \begin{tabulary}{\textwidth}{LLC}
            \toprule
            column & column & column \\
\midrule
AAA & BBB & CCC \\
\bottomrule
\end{tabulary}
\end{table}
